Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-12-01 manuscript/manuscript.Rmd
update ms.

---
title             : "Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows"
shorttitle        : "A tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  Social Brain Sciences lab, Department of Humanities, Social and Political Sciences (D-GESS).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations form a cornerstone of experimental psychology, and have had a widerspread impact on our understanding of human cognition. However, the orthodox method for analysing such data -- comparing means between conditions -- is known to conceal valuable information about the timeline of psychological effects, such as their onset time and duration. The ability to reveal finer-grained, "temporal states" of cognitive processes can have important consequences for theory development by qualitatively changing the key inferences that are drawn from psychological data. Luckily, well-established analytical approaches, such as event history analysis (EHA), are able to evaluate the detailed shape of time-to-event distributions, and thus characterise the time course of psychological states. One barrier to wider use of EHA, however, is that the analytical workflow is typically more time-consuming and complex than orthodox approaches. To help achieve broader uptake, in this paper we outline a set of tutorials that detail how to implement one distributional method known as discrete-time EHA. We illustrate how to wrangle raw data files and calculate descriptive statistics, as well as how to calculate inferential statistics via Bayesian and frequentist multilevel regression modelling. Along the way, we touch upon several key aspects of the workflow, such as how to specify regression models, the implications for experimental design, as well as how to manage inter-individual differences. We finish the article by considering the benefits of the approach for understanding psychological states, as well as the limitations and future durections of this work. Finally, the project is written in R and freely available, which means the general approach can easily be adapted to other data sets, and all of the tutorials are available as .html files to widen access beyond R-users.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology"
wordcount         : "X"
bibliography      : ["extrareferences2.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, donotrepeattitle"
appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf

@@

```{r setup, include = FALSE}
# Tutorials 1 and 4
pkg1 <- c("papaja", "citr", "tidyverse", "RColorBrewer", "patchwork")

@@

# Tutorial 2
pkg2 <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel")

@@

# Tutorial 3
pkg3 <- c("lme4", "nlme")

@@

# R references
r_refs(file = "r-references.bib", append=F) # append=F prevents multiple copies of the same citation...
my_r_citation <- cite_r(file = "r-references.bib", footnote = TRUE) ## 
```

@@

```{r plot-settings}
## theme settings for ggplot
theme_set(
  theme_bw() +
    theme(text = element_text(size = 18), 
          title = element_text(size = 18),
          legend.position = "bottom")
)

## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

@@

In experimental psychology, it is standard practice to analyse reaction times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such mean-average comparisons have been the workhorse of experimental psychology over the last century, and have had a substantial impact on theory development as well as our understanding of the structure of cognition and brain function. However, differences in mean RT conceal important pieces of information, such as when an experimental effect starts, how long it lasts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhatShapingRT2016; @panisWhenDoesInhibition2022; @wolkersdorferTemporalDynamicsSequential2020]. Such information is useful not only for the interpretation of experimental effects under investigation, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].

@@

(ref:descr-fig1-caption) Means versus distributional shapes for six different simulated data set examples. The first second after stimulus onset is divided in ten bins of 100 ms. Timebin indicates the bin rank. The first bin is (0,100], the last bin is (900,1000]. For our purposes here, it is enough to know that the distributions plotted represent the probability of an event occurring in that timebin, given that it has not yet occurred. Insets show mean reaction time per condition.

@@

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict the duration of time until people died [e.g., @williammatthewmakehamLawMortalityConstruction1860].
The same logic has been applied to psychological time-to-event data, as previously demonstrated [@panisAnalyzingResponseTimes2020].
Here, in the current paper, we focus on a distributional method for time-to-event data known as discrete-time Event History Analysis (EHA), a.k.a. survival analysis, hazard analysis, duration analysis, failure-time analysis, and transition analysis [@singerAppliedLongitudinalData2003]. We hope to show the value of EHA for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience. Moreover, we provide tutorials that provide step-by-step code and instructions in the hope that we can enable others to use EHA in a more routine, efficient and effective manner.

@@

In this paper, we focus on discrete-time EHA. We first provide a brief overview of EHA to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010]. Indeed, our primary aim here is to introduce a set of tutorials, which explain **how** to do such analyses, rather than repeat in any detail **why** you may do them.

We provide six different tutorials, which are written in the R programming language and publicly available on our Github and the Open Science Framework (OSF) pages, along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply EHA to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as a .html file, so that it can be viewed by any web browser, and thus available to those that do not use R. Finally, the manuscript itself is written in R using the papaja() package [@pap2024], which makes it computationally reproducible, in terms of the underlying data and figures.

@@

In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to *timed* accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the EHA of RT data for choice RT data. In Tutorial 3a, we illustrate how to fit the same type of multilevel regression model for RT data in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

@@

2. define time point zero (e.g., target stimulus onset, fixation onset);

@@

(ref:descr-plot2-caption) Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by 44 percent of the trials, as S(t-1=5) = .44. 
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.

@@

2. RT distributions may differ from each other in multiple ways, and hazard functions allow one to capture these differences that mean-average comparisons may conceal (Townsend, 1990).

3. EHA takes account of more of the data collected in a typical speeded response experiment, by virtue of not discarding right-censored observations. Trials with longer RTs are not discarded, but instead contribute to the risk set in each time bin.

@@

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a 2-button discrimination task. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA of the RT data (hazard and survivor functions) can be extended with a discrete-time SAT analysis of the timed accuracy data. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function, which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

@@

Figure 3 shows that the early upswing in hazard is equal for both priming conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for short waiting times (< bin 6), responses always follow the prime (and not the target, as instructed). Between 500 and 600 ms the target-triggered response channel is activated and causes response competition -- ca(6) = .5 -- and a lower hazard probability in the incongruent condition. For waiting times of 600 ms or more, the hazard of response occurrence is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct.

This joint pattern of results is interesting because it can provide meaningfully different conclusions about psychological processes compared to conventional analyses, such as computing mean-average RT across trials. Mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. For instance, if mean-average RT was higher in incongruent than congruent trials, one may conclude that cognitive mechanisms that support interference control are working as expected across trials, and are indexed by each recorded response. But such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

(ref:descr-plot3-caption) Discrete-time hazard and conditional accuracy functions. Discrete time-to-event and conditional accuracy data were simulated for 2000 trials for each of two priming conditions (congruent and incongruent prime stimuli).

@@

Descriptive, data science workflow.
We perform data wrangling following tidyverse principles and a functional programming approach [@wickhamDataScienceImport2023].
Functional programming basically means you don't write your own loops but instead use functions that have been built and tested by others.
[[more here, as necessary]].

@@

We used `r my_r_citation$r` for all reported analyses. The content of the tutorials, in terms of EHA and multi-level regression modelling, is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2018, @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

`r my_r_citation$pkgs`

<!-- This creates the footnote -->

@@

Our data wrangling procedures serve two related purposes. First, we want to summarise and visualise descriptive statistics that relate to our main research questions about the time course of psychological processes. Second, we want to produce two different data sets that can each be submitted to different types of inferential modelling approaches. The two types of data structure we label as 'person-trial' data and 'person-trial-bin' data. The 'person-trial' data (Table 1) will be familiar to most researchers who record behavioural responses from participants, as it represents the measured RT and accuracy per trial within an experiment. This data set is used when fitting conditional accuracy models (Tutorials 2b and 3b).

@@

To illustrate how to quickly set up life tables for calculating the descriptive statistics (functions of discrete time), we use a published data set on masked response priming from @panisWhatShapingRT2016. A life table includes for each time bin, the risk set (i.e., the number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), P(t), possibly ca(t), and their estimated standard errors (se). At time point zero, no events can occur and therefore h(t) and ca(t) are undefined.
In their first experiment, @panisWhatShapingRT2016 presented a double arrow for 94 ms that pointed left or right as the target stimulus with an onset at time point zero in each trial. Participants had to indicate the direction in which the double arrow pointed using their corresponding index finger, within 800 ms after target onset. Response time and accuracy were recorded on each trial. Prime type (blank, congruent, incongruent) and mask type were manipulated. Here we focus on the subset of trials in which no mask was presented. The 13-ms prime stimulus was a double arrow presented 187 ms before target onset in the congruent (same direction as target) and incongruent (opposite direction as target) prime conditions.

@@

Table 3 shows the life table for condition "blank" (no prime stimulus presented) for participant 6. A life table includes for each time bin, the risk set (i.e., the number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), P(t), possibly ca(t), and their estimated standard errors (se). At time point zero, no events can occur and therefore h(t) and ca(t) are undefined.

@@

@@

(ref:life-note-caption) The column named "bin" indicates the endpoint of each time bin (in ms), and includes time point zero. For example the first bin is (0,40] with the starting point excluded and the endpoint included. se = standard error. ca = conditional accuracy.

(ref:descr-plot4-caption) Estimated discrete-time hazard, survivor, probability mass, and conditional accuracy functions for participant 6, as a function of the passage of discrete waiting time.

@@

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data [@singerAppliedLongitudinalData2003]. Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a discrete or continuous predictor. And when you treat time as a discrete predictor, you can choose between reference coding and index coding.  
In the case of a large-*N* design without repeated measurements, the parameters of a discrete-time hazard model can be estimated using standard logistic regression software after expanding the typical person-trial data set into a person-trial-bin data set [@allisonSurvivalAnalysisUsing2010]. When there is clustering in the data, as in the case of a small-*N* design with repeated measurements, the parameters of a discrete-time hazard model can be estimated using population-averaged methods (e.g., Generalized Estimating Equations), and Bayesian or frequentist generalized linear mixed models [@allisonSurvivalAnalysisUsing2010].

@@

ptb_data <- ptb_data %>% 
# select analysis time range: (200,600] with 10 bins (time bin ranks 6 to 15)
filter(period > 5) %>%
       # continuous predictor for TIME named "period_9", centered on bin 9
mutate(period_9 = period - 9,
       # categorical predictor for TIME named "timebin" with index coding
       timebin = factor(period,levels=c(6:15)),
       # factor "condition" using reference coding, with "blank" as the reference level
       condition = factor(condition, labels = c("blank", "congruent","incongruent")),
       # categorical predictor "prime" with index coding
       prime = ifelse(condition=="blank",1, ifelse(condition=="congruent",2,3)),
       prime = factor(prime,levels=c(1,2,3)))
```
\normalsize

@@

To get the posterior distribution of each model parameter given the data, we need to specify prior distributions for the model parameters which reflect our prior beliefs.
In Tutorial_2a.Rmd we perform some prior predictive checks to make sure our selected prior distributions reflect our prior beliefs.

@@

Previous research has shown that psychological effects typically change over time [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022]. In the next model, we use index coding for both TIME (variable "timebin") and the categorical predictor prime-target-congruency (variable "prime"), so that we get 30 grand intercepts, one for each combination of timebin level and prime level. Here is the model formula of this model that we call "M1i".

@@

(ref:table-props-caption) Summarizing the posterior distributions of each contrast by their proportions below and above zero.

(ref:table-props-note-caption) prop_below = proportion below zero: prop_above = proportion above zero.

@@

If we want to focus more on inter-individual differences, we can study the subject-specific differences in hazard in Figure 9. Note that three participants (1, 2, and 3) show a negative difference for the contrast "congruent minus incongruent" in bin (360,400] -- subject 2 also in bin (320,360].

@@

In this fourth tutorial, we illustrate how to fit a Bayesian multi-level regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. The general process is similar to Tutorial 2a, except that (a) we use the person-trial data set, (b) we use the logit link function, and (c) we change the priors. For illustration purposes, we only fitted the effects of model M1i (see Tutorial 2a) in the conditional accuracy model called M1i_ca.

@@

For illustration purposes, we only fitted the effects from model M1i (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. The resulting hazard model is called M1i_f.

@@

In this sixth tutorial we illustrate how to fit a multilevel regression model to the timed accuracy data in the frequentist framework, for the data set used in Tutorial 1a. For illustration purposes, we only fitted the effects from model M1i_ca (see Tutorial 2b) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. Again, the resulting conditional accuracy model M1i_ca_f did not converge and resulted in a singular fit.

>>>>

---
title             : "Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows"
shorttitle        : "A tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  Social Brain Sciences lab, Department of Humanities, Social and Political Sciences (D-GESS).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations form a cornerstone of experimental psychology, and have had a widerspread impact on our understanding of human cognition. However, the orthodox method for analysing such data -- comparing means between conditions -- is known to conceal valuable information about the timeline of psychological effects, such as their onset time and duration. The ability to reveal finer-grained, "temporal states" of cognitive processes can have important consequences for theory development by qualitatively changing the key inferences that are drawn from psychological data. Luckily, well-established analytical approaches, such as event history analysis (EHA), are able to evaluate the detailed shape of time-to-event distributions, and thus characterise the time course of psychological states. One barrier to wider use of EHA, however, is that the analytical workflow is typically more time-consuming and complex than orthodox approaches. To help achieve broader uptake, in this paper we outline a set of tutorials that detail how to implement one distributional method known as discrete-time EHA. We illustrate how to wrangle raw data files and calculate descriptive statistics, as well as how to calculate inferential statistics via Bayesian and frequentist multilevel regression modelling. Along the way, we touch upon several key aspects of the workflow, such as how to specify regression models, the implications for experimental design, as well as how to manage inter-individual differences. We finish the article by considering the benefits of the approach for understanding psychological states, as well as the limitations and future directions of this work. Finally, the project is written in R and freely available, which means the general approach can easily be adapted to other data sets, and all of the tutorials are available as .html files to widen access beyond R-users.

keywords          : "response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology"
wordcount         : "X"
bibliography      : ["extrareferences2.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, donotrepeattitle"
appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf

@@

```{r setup, include = FALSE}
pkg1 <- c("papaja", "citr", "tidyverse", "RColorBrewer", "patchwork")

@@

pkg2 <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel")

@@

pkg3 <- c("lme4", "nlme")

@@

# R references
r_refs(file = "r-references.bib", append=F) # append=F prevents multiple copies of the same citation
my_r_citation <- cite_r(file = "r-references.bib", footnote = TRUE) 
```

@@

@@

In experimental psychology, it is standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such mean-average comparisons have been the workhorse of experimental psychology over the last century, and have had a substantial impact on theory development as well as our understanding of the structure of cognition and brain function. However, differences in mean RT conceal important pieces of information, such as when an experimental effect starts, how long it lasts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhatShapingRT2016; @panisWhenDoesInhibition2022; @wolkersdorferTemporalDynamicsSequential2020]. Such information is useful not only for the interpretation of experimental effects under investigation, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].

@@

(ref:descr-fig1-caption) Means versus distributional shapes for six different simulated data set examples. The first second after stimulus onset is divided in ten bins of 100 ms. Timebin indicates the bin rank. The first bin is (0,100], the last bin is (900,1000]. For our purposes here, it is enough to know that the distributions plotted represent the probability of an event occurring in that timebin, given that it has not yet occurred. Insets show mean response time per condition.

@@

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict the duration of time until people died [e.g., @halleyVIEstimateDegrees1997; @williammatthewmakehamLawMortalityConstruction1860].
The same logic has been applied to psychological time-to-event data, as previously demonstrated [@panisAnalyzingResponseTimes2020].

@@

In this paper, we focus on a distributional method for time-to-event data known as discrete-time Event History Analysis (EHA), a.k.a. survival analysis, hazard analysis, duration analysis, failure-time analysis, and transition analysis [@singerAppliedLongitudinalData2003]. We hope to show the value of EHA for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience. Most importantly, we provide tutorials that provide step-by-step code and instructions in the hope that we can enable others to use EHA in a more routine, efficient and effective manner.

We first provide a brief overview of EHA to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010]. Indeed, our primary aim here is to introduce the set of tutorials, which explain **how** to do such analyses, rather than repeat in any detail **why** you may do them.

We provide six different tutorials, which are written in the R programming language and publicly available on our Github and the Open Science Framework (OSF) pages, along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply EHA to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as a .html file, so that it can be viewed by any web browser, and thus available to those that do not use R. Finally, the manuscript itself is written in R using the papaja package [@pap2024], which makes it computationally reproducible, in terms of the underlying data and figures.

@@

In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We also perform prior predictive checks, compare models, and interpret the plots of the predicted hazard functions for the selected model, and the posterior distributions of our contrasts of interest. In Tutorial 2b we fit Bayesian multi-level regression models to *timed* accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the EHA of RT data for choice RT data.

In Tutorial 3a, we shortly illustrate how to fit similar multilevel regression models for RT data in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

@@

2. define time point zero (e.g., target stimulus onset, fixation onset, etc.);

@@

(ref:descr-plot2-caption) Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by 44 percent of the trials, as S(t=5) = .44. 
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.

@@

2. RT distributions may differ from each other in multiple ways, and hazard functions allow one to capture these differences which mean-average comparisons may conceal (Townsend, 1990).

3. EHA takes account of more of the data collected in a typical speeded response experiment, by virtue of not discarding right-censored observations. Trials with very long RTs are not discarded, but instead contribute to the risk set in each time bin (see below).

@@

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a discrimination task. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA of the RT data (hazard and survivor functions) can be extended with a discrete-time SAT analysis of the timed accuracy data. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function, which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

@@

Figure 3 shows that the early upswing in hazard is equal for both priming conditions, and that early emitted responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for short waiting times (< bin 6), responses always follow the prime (and not the target, as instructed). During timebin 6 the target-triggered response channel is activated and causes response competition -- ca(6) = .5 -- and a lower hazard probability in the incongruent condition. For waiting times of 600 ms or more, the hazard of response occurrence is lower in incongruent compared to congruent trials, and all responses emitted in these late bins are correct.

This joint pattern of results is interesting because it can provide meaningfully different conclusions about psychological processes compared to conventional analyses, such as computing mean-average RT and accuracy across trials. Mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. For instance, if mean-average RT was higher in incongruent than congruent trials, one may conclude that cognitive mechanisms that support interference control are working as expected across trials, and are indexed by each recorded response. But such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

(ref:descr-plot3-caption) Discrete-time hazard and conditional accuracy functions. Discrete time-to-event and conditional accuracy data were simulated for 2000 trials for each of two priming conditions (congruent and incongruent prime stimuli). Bin width equals 100 ms.

@@

We perform data wrangling following tidyverse principles and a functional programming approach [@wickhamDataScienceImport2023].
Functional programming basically means you don't write your own loops but instead use functions that have been built and tested by others.
[[more here, as necessary]].

@@

We used `r my_r_citation$r` for all reported analyses. The content of the tutorials, in terms of EHA and multi-level regression modelling, is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2018, @heiss2021,  @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

`r my_r_citation$pkgs` <!-- create footnote -->

@@

Our data wrangling procedures serve two related purposes. First, we want to summarise and visualise descriptive statistics that relate to our main research questions about the time course of psychological processes, using a life table. A life table includes for each time bin, the risk set (i.e., the number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), P(t), possibly ca(t), and their estimated standard errors (se).

Second, we want to produce two different data sets that can each be submitted to different types of inferential modelling approaches. The two types of data structure we label as 'person-trial' data and 'person-trial-bin' data. The 'person-trial' data (Table 1) will be familiar to most researchers who record behavioural responses from participants, as it represents the measured RT and accuracy per trial within an experiment. This data set is used when fitting conditional accuracy models (Tutorials 2b and 3b).

@@

To illustrate how to quickly set up life tables for calculating the descriptive statistics (functions of discrete time), we use a published data set on masked response priming from @panisWhatShapingRT2016. 
In their first experiment, @panisWhatShapingRT2016 presented a double arrow for 94 ms that pointed left or right as the target stimulus with an onset at time point zero in each trial. Participants had to indicate the direction in which the double arrow pointed using their corresponding index finger, within 800 ms after target onset. Response time and accuracy were recorded on each trial. Prime type (blank, congruent, incongruent) and mask type were manipulated. Here we focus on the subset of trials in which no mask was presented. The 13-ms prime stimulus was a double arrow presented 187 ms before target onset in the congruent (same direction as target) and incongruent (opposite direction as target) prime conditions.

@@

Table 3 shows the life table for condition "blank" (no prime stimulus presented) for participant 6.

(ref:life-note-caption) The column named "bin" indicates the endpoint of each time bin (in ms), and includes time point zero. For example the first bin is (0,40] with the starting point excluded and the endpoint included. At time point zero, no events can occur and therefore h(t=0) and ca(t=0) are undefined. se = standard error. ca = conditional accuracy. NA = undefined.

@@

(ref:descr-plot4-caption) Estimated discrete-time hazard, survivor, probability mass, and conditional accuracy functions for participant 6. Vertical dotted lines indicate the estimated median RTs. Error bars represent +/- 1 Standard Error of the respective proportion.

@@

@@

There are several analytic decisions one has to make when fitting a discrete-time hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data [@singerAppliedLongitudinalData2003]. Third, one has to choose whether to treat TIME (i.e., the time bin index t) as a discrete or continuous predictor. And when you treat a variable as a discrete predictor, you can choose between reference coding and index coding.

In the case of a large-*N* design without repeated measurements, the parameters of a discrete-time hazard model can be estimated using standard logistic regression software after expanding the typical person-trial data set into a person-trial-bin data set [@allisonSurvivalAnalysisUsing2010]. When there is clustering in the data, as in the case of a small-*N* design with repeated measurements, the parameters of a discrete-time hazard model can be estimated using population-averaged methods (e.g., Generalized Estimating Equations), and Bayesian or frequentist generalized linear mixed models [@allisonSurvivalAnalysisUsing2010].

@@

ptb_data <- ptb_data %>% 
# select analysis time range: (200,600] with 10 bins (time bin ranks 6 to 15)
filter(period > 5) %>%
       # continuous predictor for TIME named "period_9", centered on bin 9
mutate(period_9 = period - 9,
       # categorical predictor for TIME named "timebin" with index coding
       timebin = factor(period, levels = c(6:15)),
       # factor "condition" using reference coding, with "blank" as the reference level
       condition = factor(condition, labels = c("blank", "congruent","incongruent")),
       # categorical predictor "prime" with index coding
       prime = ifelse(condition=="blank", 1, ifelse(condition=="congruent", 2, 3)),
       prime = factor(prime, levels = c(1,2,3)))
```
\normalsize

@@

To get the posterior distribution of each model parameter given the data, we need to specify prior distributions for the model parameters which reflect our prior beliefs.
In Tutorial_2a.Rmd we perform a few prior predictive checks to make sure our selected prior distributions reflect our prior beliefs.

@@

Previous research has shown that psychological effects typically change over time [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhenDoesInhibition2022]. In the next model, therefore, we use index coding for both TIME (variable "timebin") and the categorical predictor prime-target-congruency (variable "prime"), so that we get 30 grand intercepts, one for each combination of timebin level and prime level. Here is the model formula of this model that we call "M1i".

@@

(ref:table-props-caption) Summarizing the posterior distributions of each contrast by their proportion below and above zero.

(ref:table-props-note-caption) prop_below = proportion below zero; prop_above = proportion above zero.

@@

If we want to focus more on inter-individual differences, we can study the subject-specific hazard functions in Figure 9. Note that three participants (1, 2, and 3) show a negative difference for the contrast "congruent minus incongruent" in bin (360,400] -- subject 2 also in bin (320,360].

@@

In this fourth tutorial, we illustrate how to fit a Bayesian multi-level regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. The general process is similar to Tutorial 2a, except that (a) we use the person-trial data set, (b) we use the logit link function, and (c) we change the priors. To keep the tutorial short, we only fitted the effects of model M1i (see Tutorial 2a) in the conditional accuracy model called M1i_ca.

@@

To keep this tutorial short, we only fitted the effects from model M1i (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. The resulting hazard model is called M1i_f.

@@

In this sixth tutorial we illustrate how to fit a multilevel regression model to the timed accuracy data in the frequentist framework, for the data set used in Tutorial 1a. To keep it short, we only fitted the effects from model M1i_ca (see Tutorial 2b) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. Again, the resulting conditional accuracy model M1i_ca_f did not converge and resulted in a singular fit.
